
% </A>
% </A>
% </A>
\week{October 16, 2000 }

Like lots of mathematicians these days, I'm trying to understand
M-theory.  It's a bit difficult, partially because the theory doesn't
really \emph{exist} yet.  If it existed, it would explain lots of stuff: on
that everyone agrees.  But nobody knows how to formulate M-theory in a
precise way, so you can't open up a paper and stare at "the fundamental
equation of M-theory", or anything like that.  There are some conjectures 
about what M-theory might be like, but no solid agreement.

One thing that \emph{does} exist is 11-dimensional supergravity.  This is 
supposed to be some kind of classical limit of M-theory.  But the good 
thing is, it's a classical field theory with a Lagrangian that you can 
write down and ponder to your heart's content.  So I'm trying to learn a 
bit about this. 
 
Unfortunately, being a mathematician, I like to understand everything 
rather carefully, preferably in a conceptual way that doesn't involve 
big equations with indices dangling all over the place.  This is slowing  
me down, because all the descriptions I've seen make 11-dimensional 
supergravity look sort of ugly, when in fact it should be really pretty. 
The physicists always point out that it's a lot simpler than the 
supergravity theories in lower dimensions.  On that I agree!  But I  
don't find it to be quite as simple as I'd like.   
 
Now, mathematicians always whine like this when they are trying to learn
physics that hasn't been pre-processed by some other mathematician.  So
just to show that I'm not completely making this stuff up, let me show
you the Lagrangian for 11d supergravity, as taken from the famous string
theory text by Green, Schwarz and Witten (see "<A HREF =
"week118.html">week118</A>"):
 
$$
L = - (1/2k^{2}) e R  
    - (1/2) e \psi *_{M} \Gamma ^{MNP} D_{N}[\omega  + \omega ')/2] \psi _{P} 
    - (1/48) e F^{2}
    - (\sqrt 2k/384) e (\psi *_{M} \Gamma ^{MNPQRS} \psi _{S} + 12 \psi *^{N} \Gamma ^{PQ} \psi ^{R}) (F + F')_{NPQR} 
    - (\sqrt 2k/3456) \epsilon ^{M1 ... M11} F_{M1 ... M4} F_{M5 ... M8} A_{M9 ... M11}
$$
    
For comparison, here's the Lagrangian for ordinary gravity: 
 
\begin{verbatim}
L = e R 
\end{verbatim}
    
Here e is the volume form and R is the Ricci scalar curvature.  Of
course, there is a lot of stuff packed into this "R".  General relativity 
didn't look so slick when Einstein first made it up!  But by now,
mathematicians have gnawed away at it for long enough that there's a
nice theory of differential geometry, where after a few months of work
you learn about "R".  And after you've done this work, you realize that
"R" is a very natural concept.  I want to get to this point for the
Lagrangian for 11d supergravity, but I'm not there yet.
 
You'll note that apart from a constant, the Lagrangian for 11d 
supergravity starts out basically like the Lagrangian for ordinary 
gravity.  So \emph{that} part I understand.  It's just the other stuff  
that's the problem. 
 
Modulo some subtleties discussed below, the whole Lagrangian is built 
from just three ingredients, which are the three basic fields in the 
theory: 
 
A) a Lorentzian metric g on the 11-dimensional manifold 
representing spacetime, 
 
B) a field \psi  on this manifold which takes values in the real  
spin-3/2 representation of SO(10,1), 
 
C) a 3-form field A on this manifold. 
 
Physicists call the metric the "graviton".  They call the spin-3/2 field 
the "gravitino" or a "Rarita-Schwinger field".  And they call the 3-form  
a "gauge field", by analogy to the 1-form that appears in electromagnetism. 
Above it's written as "A", to remind us of this analogy, but people often  
use a "C" instead - for reasons I'll explain later. 
 
Let me say a bit more about these three items.  To define a spin-3/2
field on a manifold we need to give the manifold a spin structure.
Locally, we can do this by picking a smoothly varying basis of tangent
vectors.  Such a thing is called a "frame field", but it also has other
names: in 4-dimensional spacetime people call it a "tetrad" or
"vierbein", after the German word for "four legs", but in 11-dimensional
spacetime people call it an "elfbein", after the German word for "eleven
legs".  Anyway, this frame field determines a spin structure, and also
a metric, if we declare the basis to be orthonormal.
 
The metric, in turn, determines the Levi-Civita connection on the 
tangent bundle.  However, in modern Lagrangians for gravity, people 
often treat the frame field and connection as independent variables.   
This amounts to dropping the requirement that the connection be 
torsion-free (while still requiring that it be metric-preserving).  Only 
when you work out the equations of motion from the Lagrangian do you get 
back the equation saying the connection is torsion-free - and even this 
only happens when there are no fields with \emph{spin} around.   In these 
theories, spin creates torsion!  But the torsion doesn't propagate:  it 
just sits there, determined by other fields.  So we are basically just 
repackaging the same data when we work with a frame field and connection 
instead of a metric.   
 
As a slight variant, instead of working with a frame field and 
connection on the tangent bundle, we can work with a frame field and 
"spin connection" - a connection on the spin bundle.  We need to do this 
whenever we have fields with half-integer spin around, as in supergravity.    
 
Okay, so we'll use a frame field and spin connection to describe the
graviton.  What about the gravitino?  I'm less clear about this, but I
guess the idea is that we think of the spin-3/2 representation of the
Lorentz group SO(10,1) as sitting inside the tensor product of the
spin-1 representation and the spin-1/2 representation.  This allows us
to think of the gravitino as a spinor-valued 1-form on spacetime.
That's why people write it as \psi _{N}: 
the subscript indicates that we've
got some sort of 1-form on our hands.  One thing I don't understand is
what, if any, constraints there are on a spinor-valued 1-form to make
it lie in the spin-3/2 representation.
  
What are spinors like in 11-dimensional spacetime?  For this, go back
and reread "<A HREF = "week93.html">week93</A>".  You'll see
that by Bott periodicity, spinors in (n+8)-dimensional spacetime are
just like spinors in n-dimensional spacetime, but tensored with
R^{16}.  So spinors in 11-dimensional spacetime are a lot like
spinors in 3-dimensional spacetime!  In 3 dimensions, the double cover
of the Lorentz group is just SL(2,R), and its spinor representation is
R^{2}.  Actually these are "real" spinors, or what
physicists call "Majorana" spinors.  We could complexify and
get "complex" or "Dirac" spinors - but we won't!
 
Since the space of Majorana spinors in 3d spacetime is R^{2},
the space of Majorana spinors in 11d spacetime is R^{2} tensor
R^{16} = R^{32}.  The gravitino is a 1-form taking
values in this space.
 
Finally, what about the 3-form that appears in 11d supergravity?  Why is
it called a "gauge field"?  Well, if you've made it this far,
you probably know that the 1-form in electromagnetism (the "vector
potential") is perfectly suited for integrating along the worldline
of a charged point particle.  Classically, the resulting number is just
the \emph{action}  In quantum theory, the exponential of the action describes
how the particle's \emph{phase} changes.
 
If we're dealing with strings instead of point particles, we can pull
the same trick using a 2-form, which is the right sort of thing to
integrate over the 2-dimensional worldsheet of a string.  Since people
call the 1-form in electromagnetism A, they naturally took to calling
this 2-form B.  People like to study strings propagating in a background
metric that satisfies the vacuum Einstein equations, but they also study
what happens when you throw in a background B field like this, and add a
term to the string action that's proportional to the integral of B over
the string worldsheet.  It works out nice when the B field satisfies the
obvious analogues of the vacuum Maxwell equations:
 
\begin{verbatim}
dF = 0, d*F = 0 
\end{verbatim}
    
where the "curvature" or "field strength tensor" F
is given by F = dB.
 
Like Maxwell's equations, these equations are
"gauge-invariant", in the sense that we can change B like this
without changing the field strength tensor:
 
$$
B \to  B + dw, 
$$
    
where w is any 1-form.    
 
Similarly, people believe that M-theory involves 2-dimensional membranes
called "2-branes".  A 2-brane traces out a 3-dimensional
"world-volume" in spacetime.  The 3-form field in 11d
supergravity is perfectly suited for integrating over this world-volume!
So we're really dealing with a still higher-dimensional analog of
electromagnetism.  Since we've already talked about a 1-form A that
couples to point particles and a 2-form field B that couples to strings,
it makes sense to call this 3-form C.  Lots of people do that.  But I'll
stick with Green, Schwarz and Witten, and call it A.  I'll write F for
the corresponding field strength (which is 6dA if we use their nutty
normalization).
 
Let's look at that Lagrangian again, and see how much of it we can 
understand now: 
$$
L = - (1/2k^{2}) e R  
    - (1/2) e \psi *_{M} \Gamma ^{MNP} D_{N}[\omega  + \omega ')/2] \psi _{P} 
    - (1/48) e F^{2}
    - (\sqrt 2k/384) e (\psi *_{M} \Gamma ^{MNPQRS} \psi _{S} + 12 \psi *^{N} \Gamma ^{PQ} \psi ^{R}) (F + F')_{NPQR} 
    - (\sqrt 2k/3456) \epsilon ^{M1 ... M11} F_{M1 ... M4} F_{M5 ... M8} A_{M9 ... M11}
$$
    

The number "k" is just a coupling constant.  The quantity
"e" is the volume form cooked up from the frame field.  The
quantity "R" is the Ricci scalar cooked up from the spin
connection.  "\psi _{N}" is the gravitino field, and
physicists write the inner product on spinors as
"\psi *_{N} \psi ^{N}", where the
star should really be a bar: a line drawn over the letter \psi .
"A" is the 3-form field and "F" is the field
strength.  There's also some other weird stuff I haven't explained yet.
 
Note: the first, middle, and last terms in this Lagrangian only involve 
the bosonic fields - not the gravitino.   They have the following meanings: 
 
The first term, the "e R" part, is just the Lagrangian for the 
gravitational field. 
 
The middle term is, up to a constant, just what I'd call "F ^
*F": the Lagrangian for the 3-form analog of Maxwell's equations.
 
The last term is, again up to a constant, just what I'd "F ^ F ^
A".  This is an 11-dimensional analog of the Chern-Simons term F ^
A that you can add on to the electromagnetic Lagrangian in 3d spacetime.
 
The other two terms involve the gravitino.  This is where I start getting 
nervous.  We've got this: 
 
$$
    - (1/2) e \psi *_{M} \Gamma ^{MNP} D_{N}[(\omega  + \omega ')/2] \psi _{P}
 
$$
    
and this: 
 
$$
     - (\sqrt 2k/384) e (\psi *_{M} \Gamma ^{MNPQRS} \psi _{S} + 12 \psi *^{N} \Gamma ^{PQ} \psi ^{R}) (F + F')_{NPQR} 
$$
    
The first one is mainly about how the gravitino propagates in a given 
metric - it's a kind of spin-3/2 analog of the Lagrangian for the Dirac 
equation.  The second one is mainly about the coupling of the gravitino 
to the 3-form field A - it's sort of like the coupling between the electron 
and electromagnetic field in QED.  But there's some funky stuff going on 
here!   
 
The "\Gamma " gadgets are antisymmetrized products of \gamma 
matrices, i.e. Clifford algebra generators.  I don't mind that.  It's
the stuff involving \omega ' and F' that confuses me.  "\omega "
is just a name for the spin connection, so D_{v}[\omega ] would mean
"covariant differentiation with respect to the spin
connection".  But instead of using that, we use D_{v}[(\omega  +
\omega ')/2], where \omega ' is the "supercovariantization" of the
spin connection.  Don't ask me that that means!  I know it amounts to
adding some terms that are quadratic in the gravitino field, and I know
it's required to get the whole Lagrangian to be invariant under a
"supersymmetry transformation", which mixes up the gravitino
field with the graviton and 3-form fields.  But I don't really
understand the geometrical meaning of what's going on, especially
because the supersymmetry only works "on shell" - i.e.,
assuming the equations of motion.  Similarly, I guess F' is some sort of
"supercovariantization" of the field strength tensor - but
again, it seems fairly mysterious.
 
Anyway, we can summarize all this by saying we've got gravity, a
gravitino, and a 3-form gauge field interacting in a manner vaguely
reminiscent of how gravity, the electron and the electromagnetic field
interact in the Einstein-Dirac-Maxwell equations - except that there's a
"four-fermion" term where four gravitinos interact directly.

Stepping back a bit, one is tempted to ask: what exactly is so great 
about this theory?   
 
There are various ways to focus this question a bit.  For example: the 
Lagrangian for ordinary gravity makes sense in a spacetime of any 
dimension.  The 11d supergravity Lagrangian, on the other hand, only 
makes sense in 11 dimensions.   Why is that?   
 
Well, if you ask a physicist, they'll tell you something like this: 
 
\begin{quote}
    Eleven is the maximum spacetime dimension in which one can formulate 
    a consistent supergravity, as was first recognized by Nahm in his 
    classification of supersymmetry algebras.  The easiest way to see this 
    is to start in four dimensions and note that one supersymmetry relates 
    states differing by one half unit of helicity.  If we now make the  
    reasonable assumption that there be no massless particles with spins 
    greater than two, then we can allow up to a maximum of N = 8  
    supersymmetries taking us from the helicity -2 through to helicity +2. 
    Since the minimal supersymmetry generator is a Majorana spinor with 
    four offshell components, this means a total of 32 spinor components. 
    Now in a spacetime with D dimensions and signature (1,D-1), the 
    maximum value of D admitting a 32 component spinor is D = 11. 
\end{quote}
    

In case you're wondering, this is from the first paragraph of this book: 
 
1) The World in Eleven Dimensions: Supergravity, Supermembranes and 
M-theory, ed. M. J. Duff, Institute of Physics Publishing, Bristol, 
1999. 
 
which is a collection of the most important articles on these topics. 
It's a fun book to carry around - you can really impress people with the 
title.  But if you're a mathematician trying to decipher the above 
passage, it helps to note a few things. 
 
First, this explanation of why 11d supergravity is good boils down to 
saying that it's the biggest, baddest supergravity theory around that 
doesn't give particles of spin greater than two when we compactify the 
extra dimensions in order to get a 4d theory.   
 
Second, why is it "reasonable" to assume that there aren't massless 
particles with spin greater than two?  Because it's physics folklore 
that quantum field theories with such particles are bad, nasty and evil 
- in fact, so evil that nobody even dares explain why!  Well, actually 
there's a paper by Witten in the above book that contains references to 
papers that supposedly explain why particles of spin > 2 are bad.  It's  
an excellent paper, too: 
 
2) Edward Witten, Search for a realistic Kaluza-Klein theory, Nucl. 
Phys. B186 (1981), 412-428. 
 
Maybe someday I'll get up the nerve to read those references. 
 
Third, once we buy into this "spin > 2 bad" idea, the rest
of the argument is largely stuff about spinors and Clifford algebras.
This is easy for mathematicians to learn, at least after a little
physics jargon has been explained.  For example, a "Majorana"
spinor is just a real spinor, and "offshell components" refer
to the components of a field that are independent before you impose the
equations of motion.
 
Fourth, if you're a mathematician wondering what "supersymmetry  
algebras" are, there are places where you can start learning about 
this without needing to know lots of physics: 
 
3) Quantum Fields and Strings: A Course for Mathematicians, 2 volumes,
eds. P. Deligne, P. Etinghof, D. Freed, L. Jeffrey, D. Kazhdan, D. Morrison 
and E. Witten, American Mathematical Society, Providence, Rhode Island,
1999.
 
Unfortunately, this book does not cover supergravity theories. 
 
Fifth, Nahm's classification of supersymmetry algebras looks like the 
sort of thing an algebraist should be able to understand, though I 
haven't yet understood it.  You can find it in Duff's book, or in the 
original paper:  
 
4) W. Nahm, Supersymmetries and their representations, Nucl. Phys. 
B135 (1978), 149-166. 
 
Next I want to mention some wild guesses and speculations about 11d  
supergravity and M-theory.  I'm guessing these theories are somehow 
a cousin of 3d Chern-Simons theory, related in a way that involves 
Bott periodicity.  And I'm guessing that there's something deeply 
octonionic about this theory.  There's probably something wrong about
these guesses, since I can't quite get everything to fall in line.
But there's also probably something right about them.
 
We've seen two clues already: 
 
First, the 11d spinors are related to 3d spinors via Bott periodicity,
which amounts to tensoring with R^{16} - the space of Majorana spinors
in 8d Euclidean space.  Given the relation between octonion, 8d spinors
and Bott periodicity (see "<A HREF = "week61.html">week61</A>"
and "<A HREF = "week105.html">week105</A>"), it's also very
natural to think of these Majorana spinors as pairs of octonions.
 
Second, the Chern-Simons-like term F ^ F ^ A in 11d supergravity is akin
to the 3d Chern-Simons Lagrangian F ^ A.  But this relation is a bit
odd, since a crucial part of it involves switching from a 1-form gauge
field in the 3d case to a 3-form gauge field in the 11d case.  To really
understand this, we first need to understand the geometry of these
generalized "gauge fields".  These higher gauge fields are
really not connections on bundles, but connections on
"n-gerbes", which are categorified analogues of bundles.  I
explained this to some extent in "<A HREF =
"week25.html">week25</A>" and "<A HREF =
"week151.html">week151</A>", but the basic idea is that there's an
analogy like this:
 
\begin{verbatim}
1-forms   connections on bundles     parallel transport of point particles 
2-forms   connections on gerbes      parallel transport of strings 
3-forms   connections on 2-gerbes    parallel transport of 2-branes 
4-forms   connections on 3-gerbes    parallel transport of 3-branes 
   .                 .                             . 
   .                 .                             . 
   .                 .                             . 
 
\end{verbatim}
    
and so on.  Just as connections on bundles naturally give rise to 
Chern classes and the Chern-Simons secondary characteristic classes, 
the same should be true for these higher analogues of connections. 
 
There is also another clue: as I mentioned in "<A HREF =
"week118.html">week118</A>", you can only write down Lagrangians
for supersymmetric membranes in certain dimensions.  There are
supposedly 4 basic cases, which correspond to the 4 normed division
algebras:

\begin{verbatim}
the 2-brane in dimension 4  - real numbers
the 3-brane in dimension 6  - complex numbers
the 5-brane in dimension 10 - quaternions
the 2-brane in dimension 11 - octonions
\end{verbatim}
    
Part of the point is that the in these theories there are 1, 2, 4, 
or 8 dimensions transverse to the worldvolume of the brane in question.
So 2-branes in 11 dimensions, in particular, are inherently "octonionic". 
This seems like a wonderful clue, but so far I don't really understand it.   
The evidence is lurking here: 
 
5) T. Kugo and P. Townsend, Supersymmetry and the division algebras, 
Nucl. Phys. B221 (1983), 357-380. 
 
6) G. Sierra, An application of the theories of Jordan algebras and 
Freudenthal triple systems to particles and strings, Class. Quant.  
Grav. 4 (1987) 227. 
 
7) J. M. Evans, Supersymmetric Yang-Mills theories and division algebras, 
Nucl. Phys. B298 (1988), 92. 
 
8) M. J. Duff, Supermembranes: the first fifteen weeks, Class. Quant. 
Grav. 5 (1988), 189-205. 
 
There are also tantalizing clues scattered through these fascinating 
books: 
 
9) Feza Gursey and Chia-Hsiung Tze, On the Role of Division, Jordan, and 
Related Algebras in Particle Physics, World Scientific, Singapore, 1996. 
 
10) Jaak Lohmus, Eugene Paal and Leo Sorgsepp, Nonassociative Algebras 
in Physics, Hadronic Press, Palm Harbor, Florida, 1994. 
 
However, these books are frustrating to me, because they make some 
interesting claims without providing solid evidence.   
 
Anyway, I'll try to keep gnawing away at this bone until I get to the 
marrow!   Any help would be appreciated.
 

\par\noindent\rule{\textwidth}{0.4pt}
\textbf{Addenda:}
Here is an article that Maxime Bagnoud posted to sci.physics.research,
which answers some of my questions above....



% parser failed at source line 518
